Using Cell Phone Keyboards is (\NP) Hard

### 2010-01-20 article.tex
Woot!  Proof going in!

\newcommand{\newword}[1]{{\bf #1}}
\newcommand{\NP}{\ensuremath{\mathcal{NP}}}
\newcommand{\Instance}{\Instance}
\newcommand{\Question}{~\\
\Question}
\spnewtheorem{prob}[theorem]{Problem}{\bfseries}{\itshape}

@@

\begin{prob}[{\sc BasicCellPhoneTyping}]~\\
\Instance\ A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$. \\
        ~\\
\Question\ What is the best partition $P$ of $A$ into $k$ sequences, such
that the total number of button presses to type every word in $W$ its
associated frequency times is minimized?  The number of button presses is equal
to the order of the letter in the sequence assigned to a given key.  For
example, in the key $2\to\{abc\}$, $a$ requires one keystroke and $c$ requires
3.  Equivalently, if we denote the number of button presses required to type
letter $a$ with partition $P$ as $\mathrm{IN}_P(a)$, then we want to find the $P$ that
minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} \mathrm{IN}_P(c) * f$$
\label{bcpt}
\end{prob}

@@

By an argument symmetric to the proof of Theorem~\ref{basicthm}, we find that
the optimal layout in this new scheme is to place the letters of a given key in
sorted order.  Thus, the keyboard layout changes to $2\to[acb], 3\to[edf],
4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv], 9\to[wyxz]$ and
requires 678,547,463 button presses to enter the whole corpus.  This represents
a $28.43\%$ savings, but has the added benefit that it does not change which
key is mapped to which set of letters.  This makes it \newword{legacy
preserving}, as creating a keyboard with this layout will not invalidate such
advertising gems as 1-800-FLOWERS.  Thus, we can speed up users by $28.43\%$
(neglecting inter-letter pauses) and not undermine 50 years of advertising.
This layout represents perhaps the most plausible layout yet\footnote{The
described layout is the only legacy preserving layout in this paper, and
therefore should be considered the most practical suggestion, despite the fact
that its algorithm is among the least FUN.  It also has the added benefit of
not angering any of the organizations behind the numbers 1-800-FLOWERS,
1-800-THRIFTY, and 1-800-MARINES}.  After setting up this baseline for the easy
problem, we turn our attention to optimizing cellphone keyboards for predictive
input methods.

@@

In an effort to minimize the pain of cellphone keyboard typing, cellular
telephone manufacturers have created the T9 input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide \newword{speculative
lookahead} to report, at any given time, what word it is that the user is most
likely trying to enter and provide a completion.  Unfortunately, because there
are fewer keys on the cellphone keyboard than there are letters in the English alphabet, there are words
which have different spellings but the same input sequence.  As an example, in
the traditional mapping of keys to characters
both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same
input sequence force the user to press a third button to cycle through the
possibilities in order from most likely (``of'') to least likely (``me'').
Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and
``goof'' all have the input sequence ``4663''.  If the * key is used as the
``next match'' button, then the user will actually have to type in
``4663******'' to type in the word ``goof'', for a total of 10 button presses --- exactly the same number of button presses required using the default layout and the basic input method, and three more button presses than is required when using the basic input method with an optimized layout.

@@

\begin{prob}[{\sc MinimumT9Keystrokes}]~\\
\label{thm:minstrokes}
\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).\\
~\\
\Question\ What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
\end{prob}

@@

\begin{prob}~\\
\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.\\
~\\
\Question\ Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
\end{prob}

Before proceeding, we note that this we note that this problem is in \NP, as
any assignment may be verified to have and expected number of characters of
less than $f$ in time proportional to the total length of all the words in $W$.
In order to prove completeness, however, we first prove the \NP-completeness of an intermediate problem: {\sc UniquePathColoring}.

\begin{prob}[{\sc UniquePathColoring}]~\\
\label{upcolor}
\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.\\
~\\
\Question\
\end{prob}

>>>>

\newcommand{\newword}[1]{{\bf #1}}
\newcommand{\NP}{\ensuremath{\mathcal{NP}}}
\newcommand{\Instance}{{\sc Instance~}}
\newcommand{\Question}{~\\
{\sc Question~}}
\spnewtheorem{prob}[theorem]{Problem}{\bfseries}{\itshape}

@@

\begin{prob}[{\sc BasicCellPhoneTyping}]~\\
\Instance\ A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$.

\Question\ What is the best partition $P$ of $A$ into $k$ sequences, such
that the total number of button presses to type every word in $W$ its
associated frequency times is minimized?  The number of button presses is equal
to the order of the letter in the sequence assigned to a given key.  For
example, in the key $2\to\{abc\}$, $a$ requires one keystroke and $c$ requires
3.  Equivalently, if we denote the number of button presses required to type
letter $a$ with partition $P$ as $\mathrm{IN}_P(a)$, then we want to find the $P$ that
minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} \mathrm{IN}_P(c) * f$$
\label{bcpt}
\end{prob}

@@

By an argument symmetric to the proof of Theorem~\ref{basicthm}, we find that
the optimal layout in this new scheme is to place the letters of a given key in
sorted order.  Thus, the keyboard layout changes to $2\to[acb], 3\to[edf],
4\to[ihg], 5\to[lkj], 6\to[onm], 7\to[srpq], 8\to[tuv], 9\to[wyxz]$ and
requires 678,547,463 button presses to enter the whole corpus.  This represents
a $28.43\%$ savings, but has the added benefit that it does not change which
key is mapped to which set of letters.  This makes it \newword{legacy
preserving}, as creating a keyboard with this layout will not invalidate such
advertising gems as 1-800-FLOWERS.  Thus, we can speed up users by $28.43\%$
(neglecting inter-letter pauses) and not undermine 50 years of advertising.
This layout represents perhaps the most plausible layout yet\footnote{The
described layout is the only legacy preserving layout in this paper, and
therefore should be considered the most practical suggestion.  It also has the
added benefit of not angering any of the organizations behind the numbers
1-800-FLOWERS, 1-800-THRIFTY, and 1-800-MARINES}.  After setting up this
baseline for the easy problem, we turn our attention to optimizing cellphone
keyboards for predictive input methods.

@@

In an effort to minimize the pain of cellphone keyboard typing, cellular
telephone manufacturers have created the T9 input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide
\newword{speculative lookahead} to report, at any given time, what word it is
that the user is most likely trying to enter and provide a completion.
Unfortunately, because there are fewer keys on the cellphone keyboard than
there are letters in the English alphabet, there are words which have different
spellings but the same input sequence.  As an example, in the traditional
mapping of keys to characters both ``me'' and ``of'' have the input sequence
``63''.  Two words with the same input sequence force the user to press a third
button to cycle through the possibilities in order from most likely (``of'') to
least likely (``me'').  Even worse: ``home'', ``good'', ``gone'', ``hood'',
``hoof'', ``hone'', and ``goof'' all have the input sequence ``4663''.  If
the * key is used as the ``next match'' button, then the user will actually
have to type in ``4663******'' to type in the word ``goof'', for a total of 10
button presses --- exactly the same number of button presses required using the
default layout and the basic input method, and three more button presses than
is required when using the basic input method with an optimized layout.

@@

\begin{prob}[{\sc MinimumT9Keystrokes}]~\\
\label{thm:minstrokes}
\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).

\Question\ What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
\end{prob}

@@

\begin{prob}~\\
\Instance\ An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.

\Question\ Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
\end{prob}

Note that this problem is in \NP, as any assignment may be verified to require
fewer than $f$ button pushes in time proportional to the
total length of all the words in $W$.  In order to prove completeness, however,
we first prove the \NP-completeness of an intermediate problem: {\sc
UniquePathColoring}.

\begin{prob}[{\sc UniquePathColoring}]~\\
\label{upcolor}
\Instance\ A graph $G=(V,E)$, a set of paths $P$, and a parameter $k$.  A path $p$ is a sequence of vertices in which adjacent vertices in $p$ are also adjacent in the edge set $E$.

\Question\ Is there a $k$-coloring of $G$ such that every path $p\in P$ has a unique coloring?  If we consider the coloring function $\chi(v)$ to map vertices to colors, then we can extend this notation by having $\chi(p)$ map a path $p = [ v_1, v_2, \ldots ]$ to the sequence $\chi(p) = [ \chi(v_1), \chi(v_2), \ldots ]$.  Is there a coloring $\chi$ of $V$ such that 
    $$|\left\{ \chi(p)~\forall p \in P\right\}| = |P|\enspace ?$$
\end{prob}

\begin{proof}[{\sc UniquePathColoring} is \NP-complete]
To prove that {\sc UniquePathColoring} is \NP-complete, we begin by noting that any coloring of $V$ may be verified, in polynomial time, to map each path to a unique sequence of colors.  Therefore, the problem is in \NP.  To prove completeness, we reduce from {\sc GraphColoring}\cite{gandj}.

An instance of {\sc GraphColoring } consists of a graph $G=(V,E)$ and a parameter $k$.  We then ask the question of whether there is a $k$-coloring $\chi$ of the vertices of the graph such that $\forall (u,v)\in E$, $\chi(u) \neq \chi(v)$.  We transform an instance of {\sc GraphColoring} into an instance of {\sc UniquePathColoring} in the following manner:

Given $G=(V,E)$ and $k$ from {\sc GraphColoring}, we create 
$$G'=\left(V \cup \{0,1\}, E \cup \{(0,1)\} \cup \{ (v, 0)~ \forall v \in V\} \right)$$
We then uniquely number each edge in $E$ with the numbers $1 \ldots |E|$.  For each edge $e = (u,v)$ numbered $i$, we create the path set 
\begin{eqnarray*}
p_i &=& \{[v,0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}], \\
    & &  ~[u, 0,b_1(i),b_2(i), b_3(i), \ldots b_{\lceil \log_2 i\rceil}]\} \\
\end{eqnarray*}
where $b_1(i)$ is the first digit of $i$ in binary, $b_2(i)$ is the second digit of $i$ in binary, and so on.  Now we create our path set 
$$P = \{ [0,1], [1,0] \} \cup \bigcup_{i = 1}^{|E|} p_i \enspace .$$
We then ask the question of whether $G'$ and $P$ can be unique-path colored using only $k$ colors.

If there is a $k$-coloring $\chi$ of $G$, then we use that coloring to generate
a $k$-unique-path coloring of $G'$ and $P$ in the following manner: First,
assign vertices $0$ and $1$ different colors from the range of $\chi$.  Next,
assign each vertex to the color it received in the $k$-coloring of $G'$.  Now,
every element of our path set corresponds to a unique color sequence.  $[0,1]$
and $[1,0]$ are the only sequences of length two, and because we assigned these
two vertices different colors, $\chi([0,1]) \neq \chi([1,0])$.  If path $x$ and
path $y$ are not from the same $p_i$, then they have a different sequence of
zeroes and ones.  Therefore, because $0$ and $1$ are assigned different colors,
the only possible path that a given path in some $p_i$ might be confused with
is the other path in that $p_i$.  But each $p_i$ corresponds to a an edge in
$E$, and we know for all edges in $(u,v) \in E$ that $\chi(u) \neq \chi(v)$.
Therefore, both paths within a given $p_i$ also have a distinct color sequence.  Therefore, given a $k$-coloring of $G$, we can generate a $k$-unique-path coloring of $G'$ and $P$.

Now we prove that given a $k$-unique-path coloring of $G'$ and $P$ we can create a $k$-coloring of $G$.

\end{proof}
